---
title: "City Bike Data Analysis"
author: "Robin Lee"
date: "December 18, 2015"
output: pdf_document
---

## Introduction
I am studying how weather (temperature and preciptation) affects the number of trips taken in City Bike.
Riders might be less inclined to ride bicycles when there's rain or snow. Also, a sudden drop 

Hypothesis - 
Higher temperature is associated with more trips taken at a given day. 
Rain or snow would reduce the number of trips

IV
temperature
precipation (continuous/inch)
membership type
DV
Daily total of trips
Daily total time

.  Explain what topic you are studying and what question in particular you are asking; this 
require you to say what are your main independent and
dependent variable(s).  Then state your 
hypothesis about how the IV and DV should be related.
No literature review is required at all.

## Description of Data Set and Variables
 I want to know how it was collected, how is it 
representative (or not), etc.  Everyone must tell me what variables you are using and how they have been 
coded and recoded.  

I obtain my data from two sources. One source gives me city bike trip data. The other gives me daily weather data. 
City bike trip data is obtained from City Bike System Data (https://www.citibikenyc.com/system-data). The website contains trip level data for each month from July 2013 to December 2015. I analyzed the data from Jan 2014 to December 2014. Because the dataset is trip-level, I then aggregate the number of trips, total time of trips by membership for each day. 

a code that shows how I obtain the data
citybike_scrape.R

```{r}
```


this is a sample dataset for trip level data
```{r}

```

how I aggregate it into daily level
```{r}
```

this is a sample dataset for daily aggregate data
```{r}

```

I obtain the weather data from WeatherUnderground.com (http://www.wunderground.com/history/airport/KNYC/2014/1/1/CustomHistory.html?dayend=31&monthend=12&yearend=2014&req_city=&req_state=&req_statename=&reqdb.zip=&reqdb.magic=&reqdb.wmo=&MR=1). I look at the weather data from Jan 1, 2014 to Dec 31, 2014. 

The variables of the weather data are

Then I combined the two data sources into one dataset

variables list


## Descriptive Statistics
.  You should 
have a table with the means, st.devs. etc. of all the variables you are 
using.  With explanation and commentary.
## Initial Models
.  Tell me what model you are using and why (logit, probit, LPM, fixed effects, etc.).  Start 
off with a simple model relating yo
u main IV to your main DV.  Explain the relationship and why this 
initial model is insufficient.  Maybe you need to make a scale/index of variables.  Maybe you need to 
control for additional factors.  Maybe you want to include interaction terms. Maybe you 
need to check 
for serial correlation.  Etc.  Interpret everything correctly (ceteris paribus, on 
the 
right scale, etc.)

My initial model is a OLS model. The Y variable is the number of trips. The X variables are daily average temperature and precipation. 

```{r}


```

room for improvement 
1. change of temperature compared to the previous day: a 60 degree weather may be considered warm if the previous day's temperature is 50, but would be considered cold to the rider if the previous day's temperature is 70. 

2. membership type and interaction
Weather change might affect short-term and annual rider differently. Annual riders are likely to be commuters and would be less affected by the change in weather. Short-term riders may be tourists or people who are trying out city bike. I would expect them to be affected by the change in weather more. 

3. temperature (generalized)
I suspect there be a quadratic function

plot # of trips vs temperature


4. number of stations available
If city bike has expanded over the year, then I would expect it to be an exogenous factor not captured by the initial model. 

5. weekday vs weekend

## Final Models
.  Tell me what models are your final best ones.  Why are they better than the initial ones 
a
nd what else do we need to know
?
Were there some intermediate steps to get to this final model. 
Maybe have some diagnostics and checks.



## Conclusion
.  Tell me what you learned from doing the project.  Was your initial hypothesis supported or 
not.  Why or why not?  Tell me what else you w
ish you could have done.  Tell me how this work may 
inform your thesis.  Etc

Limitation - city bike station might increase over time. Changes in number of trips could be affected by 